---
project:
  type: website

website:
  title: "Rent vs Buy – Help Maya Decide"

format:
  html:
    theme: flatly
    css:
      - styles/latex-fonts.scss

editor: visual
---

## Lab 3 -- Buy vs Rent (10-Year Outlook)

::: callout-note
**Watch Maya's 10-second introduction** before you compare the options.
:::

```{=html}
<div style="position:relative;padding-bottom:56.25%;height:0;">
  <iframe src="https://www.youtube.com/embed/vzJ31Tz5z-s"
          title="Maya intro"
          style="position:absolute;top:0;left:0;width:100%;height:100%;border:0;"
          allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"
          allowfullscreen>
  </iframe>
</div>
```
Maya knows Austin housing is pricey, but she also knows staying on the sidelines for too long can be costly. Your job in **Lab 3** is to use the spreadsheet template from Lab 2 as a **decision tool** that answers two core questions:

1.  **Can Maya afford to buy a home today** on her \$65 000 salary---while still contributing to retirement and paying down debt?\
2.  If not, **what salary (or trade-offs) would she need** to make buying feasible?

**Bonus (+ 20 pts):** At a feasible salary, **which path leaves her wealthiest after 10 years**---buying one of the three properties or continuing to rent? This requires running the Buy vs Rent models and accounting for different financing types and the retirement opportunity cost.

You will compare four realistic paths:

-   Buying the **Barton-Springs condo** a stone's throw from work\
-   Buying a smaller **starter home on Lightfoot Trail**\
-   Buying a lower-priced **starter in Kyle, TX**, 22 miles away\
-   **Continuing to rent** a two-bed South Lamar apartment with a roommate

At the end you will recommend a path **and** explain the trade-offs of affordability, down-payment size, PMI vs FHA mortgage insurance, retirement contributions, and lifestyle factors like commute and HOA dues.

------------------------------------------------------------------------

### 1 Baseline facts you must load into the model

|                             | **Option 1 -- Barton-Springs Condo** | **Option 2 -- Lightfoot Starter** | **Option 3 -- Kyle Starter** | **Rent -- South Lamar apt**   |
|---------------|---------------|---------------|---------------|---------------|
| Price / Monthly rent        | \$649 500                            | \$500 000                         | **\$350 000**                | \$2 295 / mo (≈ \$1 148 each) |
| Beds / Baths / Sq ft        | 2 / 2 / 994                          | 3 / 2 / 1 781                     | 3 / 2 / 1 590                | 2 / 2 / 997                   |
| HOA dues (mo)               | \$485                                | \$125                             | **\$46**                     | ---                           |
| Distance to office          | 0.7 mi                               | 4.2 mi                            | **22 mi**                    | 0.9 mi                        |
| **Commute cost (mo)**       | \$0                                  | \$0                               | **\$635**                    | \$0                           |
| Property-tax rate           | 1.8 %                                | 2.0 %                             | **2.05 %**                   | ---                           |
| Maintenance allowance       | 1 % of value                         | 1 %                               | 1 %                          | ---                           |
| Home-owner's insurance (mo) | \$100                                | \$110                             | **\$105**                    | ---                           |
| Utilities (mo)              | \$180                                | \$220                             | \$220                        | \$180 (split)                 |
| Renter's insurance (mo)     | ---                                  | ---                               | ---                          | \$15 each                     |
| Home appreciation           | 2.5 % / yr (all)                     |                                   |                              |                               |
| Sale / purchase costs       | 1 % / 3 % (all buys)                 |                                   |                              | ---                           |

Maya starts on a **\$65 000 salary**, expects **3 % raises**, pays **12 % federal tax**, and invests any leftover cash at **8 % pre-tax**. Rent is assumed to rise **3 % per year**. Keep these figures editable but locked on your first submission. The standard deduction for single filers for tax year 2025 is **\$15,750**. The current SALT cap is **\$10,000** but test for the updated one under the "**One big beautiful bill Act**" **\$40,000**

```{=html}
<style>
.cards{
  display:grid;
  grid-template-columns:repeat(auto-fill,minmax(320px,1fr));
  gap:1rem;margin-top:1rem}
.card{
  border:1px solid #ddd;border-radius:8px;box-shadow:0 1px 4px rgba(0,0,0,.06);
  overflow:hidden;background:#fff;font-family:system-ui,sans-serif}
.card-header{position:relative}
.card-header img{width:100%;height:200px;object-fit:cover}
.badge{
  position:absolute;top:8px;left:8px;padding:2px 6px;border-radius:4px;
  font-size:.75rem;color:#fff;opacity:.9}
.badge.Option1 { background:#0a7d00; }
.badge.Option2 { background:#006aff; }
.badge.Option3 { background:#4d4dff; }
.badge.rentOption { background:#ff8c00; }
.card-body{padding:.75rem}
.price{margin:0;font-size:1.25rem;font-weight:700}
.meta,.address{margin:.2rem 0;color:#555;font-size:.9rem}
.btn{
  display:inline-block;margin-top:.3rem;padding:4px 8px;font-size:.8rem;
  color:#006aff;border:1px solid currentColor;border-radius:4px;text-decoration:none}
.btn:hover{background:#f0f8ff}

/* Phase callout boxes */
.phase-box {
  border-left: 4px solid;
  border-radius: 4px;
  padding: 1rem 1.25rem;
  margin: 1.5rem 0;
}
.phase-box.affordability { border-color: #c0392b; background: rgba(192,57,43,0.06); }
.phase-box.goalsalary   { border-color: #2980b9; background: rgba(41,128,185,0.06); }
.phase-box.comparison    { border-color: #27ae60; background: rgba(39,174,96,0.06); }
.phase-label {
  font-weight: 700; font-size: 0.85rem; text-transform: uppercase;
  letter-spacing: 0.05em; margin-bottom: 0.25rem;
}
.phase-box.affordability .phase-label { color: #c0392b; }
.phase-box.goalsalary   .phase-label { color: #2980b9; }
.phase-box.comparison    .phase-label { color: #27ae60; }
</style>
```
::: callout-note
These three properties have been pre-selected for Maya (distance to 2222 Barton Springs Rd and price range fit).\
Click **"View Listing"** to open the live Zillow page. **Please note that prices may have changed since the time of recording this assignment.**
:::

## Buying

```{r cards, echo=FALSE, results='asis', warning=FALSE}
library(jsonlite)
library(htmltools)

# --- read the JSON once ---
listings <- fromJSON("data/maya_listings.json", simplifyDataFrame = FALSE)

# --- helpers -------------------------------------------------
badge_html <- function(category) {
  labs <- c(Option1 = "Option 1",
            Option2 = "Option 2",
            Option3 = "Option 3")
  div(class = paste("badge", category), labs[[category]])
}

one_card <- function(h) {
  div(class = "card",
      div(class = "card-header",
          img(src = h$image_url, alt = "", loading = "lazy"),
          badge_html(h$category)
      ),
      div(class = "card-body",
          p(class = "price", sprintf("$%s", format(h$price, big.mark=","))),
          p(class = "meta",
            sprintf("%s bd | %s ba | %s sq ft",
                    h$beds, h$baths, format(h$sqft, big.mark=","))),
          p(class = "meta",
            sprintf("HOA $%s/mo • %.2f mi to office",
                    format(h$hoa_monthly, big.mark=","), h$distance_miles)),
          p(class = "address", h$address),
          a(class = "btn", href = h$listing_url, target = "_blank",
            "View Listing ↗")
      )
  )
}

# --- build the grid of cards and return it as HTML ------------
tagList(
  div(class = "cards", lapply(listings, one_card))
) |> browsable()
```

## Renting

```{r rent_card, echo=FALSE, results='asis', warning=FALSE}
library(jsonlite)
library(htmltools)

rent <- fromJSON("data/maya_rent.json", simplifyDataFrame = FALSE)

rent_card <- function(h){
  div(class = "card",
    div(class = "card-header",
        img(src = h$image_url, alt = "", loading = "lazy"),
        div(class = "badge rentOption", "Rent Option")
    ),
    div(class = "card-body",
        p(class = "price",
          sprintf("$%s / mo  (%s each)",
                  format(h$rent_total, big.mark = ","),
                  format(h$rent_per_person, big.mark = ","))),
        p(class = "meta",
          sprintf("%s bd | %s ba | %s sq ft",
                  h$beds, h$baths, format(h$sqft, big.mark=","))),
        p(class = "meta",
          sprintf("%.2f mi to office", h$distance_miles)),
        p(class = "address", h$address),
        a(class = "btn", href = h$listing_url, target = "_blank",
          "View Listing ↗")
    )
  )
}

tagList(div(class = "cards", rent_card(rent))) |> browsable()
```

------------------------------------------------------------------------

### 2 The three questions you must answer (in order)

```{=html}
<div class="phase-box affordability">
  <div class="phase-label">Phase 1 — Affordability gate</div>
  <strong>Can Maya afford to buy any of these homes today?</strong>
  <p style="margin:0.5rem 0 0">
    <strong>Where to work:</strong> Start on the <code>input</code> tab — confirm Maya's salary, retirement contribution rate 
    (cell D75), debt balances, and living expenses are loaded correctly. Then go to the 
    <code>5 year Budgeting (rent or buy)</code> tab. This sheet builds a month-by-month cash-flow projection 
    that accounts for salary, retirement contributions, taxes, debt payments, and all living expenses.
  </p>
  <p style="margin:0.5rem 0 0">
    From the budgeting tab, identify Maya's <strong>monthly savings</strong> (row 55) — this is the cash left over after 
    everything except housing. That number is what she has available to put toward housing costs each month.
    Now compute the <strong>total monthly cost</strong> of each buy option (mortgage P&I + property tax + maintenance + 
    homeowner's insurance + HOA + utilities + commute + PMI/MIP if applicable) and the rent option (her share of rent + utilities + renter's insurance).
  </p>
  <p style="margin:0.5rem 0 0">
    Apply the standard lending guidelines to each scenario:
  </p>
  <ul style="margin:0.25rem 0">
    <li><strong>28 % rule</strong> — housing expense should not exceed 28 % of gross monthly income</li>
    <li><strong>36 % rule</strong> — housing + all debt payments should not exceed 36 % of gross monthly income</li>
  </ul>
  <p style="margin:0.5rem 0 0">
    <strong>In your memo:</strong> Report the monthly cash available, the monthly cost of each option, and whether 
    each passes or fails the 28/36 % tests. What does this tell you about Maya's readiness to buy at her current salary?
  </p>
</div>

<div class="phase-box goalsalary">
  <div class="phase-label">Phase 2 — What would make buying feasible?</div>
  <strong>What salary would Maya need to afford each option?</strong>
  <p style="margin:0.5rem 0 0">
    <strong>Where to work:</strong> For each buy option you want to test, first go to the <code>input</code> tab and 
    enter the property details in the Buying section (home price, HOA, property-tax rate, utilities, insurance, etc.). 
    Then open the <code>Buying Model</code> tab — look at the <strong>implied monthly housing expense</strong> (row 25 for 
    pre-tax, row 42 for after-tax benefit). That is the true monthly cost of owning that property under that financing scenario.
  </p>
  <p style="margin:0.5rem 0 0">
    Now bring that monthly housing figure back to the <code>5 year Budgeting (rent or buy)</code> tab. 
    Replace the rent/mortgage line with the buying cost (ignore any double-accounting for utilities for now). 
    Check whether Maya's monthly savings (row 55) stays positive — if it goes deeply negative, 
    she cannot sustain that option.
  </p>
  <p style="margin:0.5rem 0 0">
    Next, adjust the salary cell (D7 on the <code>input</code> tab) upward — use Goal Seek or manually iterate — until 
    the budget is sustainable under the 28/36 % rules. Test at least two scenarios:
  </p>
  <ul style="margin:0.25rem 0">
    <li>Keeping the <strong>17 % retirement contribution</strong> (cell D75)</li>
    <li>Reducing retirement to the <strong>employer-match-only level</strong> (e.g., 3–6 %) — change cell D75 and observe 
        how the budgeting tab's savings line shifts</li>
  </ul>
  <p style="margin:0.5rem 0 0">
    Also use the <code>5 year Budgeting</code> tab to check: how much has Maya <strong>saved toward a down payment</strong> 
    after 5 years of renting (look at cumulative savings row 57 and the investment balance row 67)? 
    Is it enough for 20 % down on the cheapest option? For 3.5 % FHA?
  </p>
  <p style="margin:0.5rem 0 0">
    <strong>In your memo:</strong> Present the break-even salary for each option under both retirement scenarios. 
    Discuss the trade-off: what does Maya give up in long-term retirement wealth if she cuts contributions to afford a home sooner?
  </p>
</div>

<div class="phase-box comparison">
  <div class="phase-label">Bonus — Phase 3: Buy vs. Rent net-worth comparison (+20 pts)</div>
  <strong>At a feasible salary, which path leaves Maya wealthiest after 10 years?</strong>
  <p style="margin:0.5rem 0 0">
    <em>This phase is optional bonus credit. Complete Phases 1 and 2 first.</em>
  </p>
  <p style="margin:0.5rem 0 0">
    <strong>Where to work:</strong> Your <code>input</code> tab should already have the feasible salary and property details 
    from Phase 2. Now use the <code>Buying Model</code> and <code>Renting Model</code> tabs to read the 10-year net-worth 
    projections. Repeat for each buy option (change the property inputs each time) and record the results.
  </p>
  <p style="margin:0.5rem 0 0">
    <strong>Financing type matters here.</strong> The choice of Conventional 20 %, Conventional 5 % + PMI, or FHA 3.5 % + MIP 
    significantly changes the monthly cost, the equity build-up, and therefore the 10-year net-worth outcome. 
    Test at least two financing scenarios per property and compare.
  </p>
  <p style="margin:0.5rem 0 0">
    <strong>Retirement caveat:</strong> The Buying and Renting models assume retirement contributions stay the same 
    across both paths. If Maya had to reduce her 401(k) rate to afford buying, run the analysis twice — once with the 
    buyer's rate and once with the renter's rate — and add the retirement balance gap to the renter's net worth for a fair comparison.
  </p>
  <p style="margin:0.5rem 0 0">
    <strong>In your memo:</strong> Name the best option and support it with numbers. Show how financing type 
    (Conventional vs FHA, high vs low down payment) affects the outcome. Does the ranking change 
    when you include the retirement opportunity cost?
  </p>
</div>
```

------------------------------------------------------------------------

### 3 Financing menus

The template already includes a dropdown on the Inputs sheet:

1.  **Financing Type** → Conventional High-Down \| Conventional Low-Down + PMI \| FHA High-Down + MIP \| FHA Low-Down + MIP\
2.  **Down-payment %** → automatically looked up from the table below based on your financing choice

The formulas react instantly to these choices. The table below shows the default rates and insurance rules built into the spreadsheet. Students may change the down-payment percentage to test other scenarios, but the memo must cite the default values.

| Path                                         | Down-payment default | Rate (30-yr fixed) | Up-front fee                 | Monthly insurance rule                                                                                   |
|---------------|---------------|---------------|---------------|---------------|
| **Conventional -- High-Down ≥ 25 %**         | 25 %                 | **6.50 %**         | none                         | **none**                                                                                                 |
| **Conventional -- Low-Down + PMI (\< 20 %)** | 4.5 %                | **6.60 %**         | none                         | **PMI = 0.65 % / yr × loan balance**<br>• Paid monthly<br>• Auto-cancels at **78 % LTV**                 |
| **FHA -- High-Down \> 10 % + MIP**           | 10 %                 | **6.25 %**         | **1.75 %** of loan, financed | **MIP = 0.55 % / yr × loan balance**<br>• Drops after **11 yrs** or when LTV ≤ 78 %                      |
| **FHA -- Low-Down \< 10 % + MIP**            | 4 %                  | **6.25 %**         | **1.75 %** of loan, financed | **MIP = 0.50 % / yr × loan balance**<br>• Paid monthly<br>• Lasts **full 30 yrs** if original DP \< 10 % |

------------------------------------------------------------------------

### 4 Deliverables & grading

**What to turn in**

| Item                    | Details                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                               |
|-------------|-----------------------------------------------------------|
| **Updated spreadsheet** | Use the template provided on Canvas --- **do not create a new workbook**. Modify the input cells only. Save as **`Lab3_Rent-vs-Buy_yourname.xlsx`**.                                                                                                                                                                                                                                                                                                                                                                                                                                                  |
| **One-page memo (PDF)** | Your memo must address Phases 1 and 2:<br>• **Phase 1:** State whether Maya can afford to buy at her current salary. Show the 28/36 % test results.<br>• **Phase 2:** Report the salary she would need for each option, with and without full retirement contributions. Discuss the retirement trade-off.<br>• Explain the role of down-payment size, PMI / MIP drag, commute cost, and risk factors (HOA hikes, etc.).<br>• **Bonus (Phase 3):** If attempted, name the best buy-vs-rent option at a feasible salary and show how financing type and retirement opportunity cost affect the ranking. |

> **Reminder:** I will grade primarily on the quality of your reasoning, but I will double-check your numbers --- be sure they reconcile with the Inputs you chose.

**Grading rubric (100 pts + up to 20 bonus)**

| Category                                                                                                                 | Pts          |
|----------------------------------------------------------|--------------|
| **Phase 1:** Correct affordability analysis with 28/36 % tests                                                           | **20**       |
| **Phase 2:** Salary goal-seek and retirement trade-off analysis                                                          | **30**       |
| Spreadsheet inputs adjusted correctly & formulas intact                                                                  | **10**       |
| Memo: clear recommendation with numeric support for Phases 1 & 2                                                         | **40**       |
| **Bonus --- Phase 3:** 10-year buy-vs-rent net-worth comparison with financing scenarios and retirement opportunity cost | **up to 20** |

Submit both files to Canvas by **Friday 20th March, 11:59 pm**.\
*No late work accepted.* Good luck!

------------------------------------------------------------------------

### Appendix -- Modelling assumptions

These assumptions apply to the spreadsheet model. You will need them primarily for Phase 2 and the bonus Phase 3.

#### Additional modelling assumptions

| Item                                             | Value / Rule                                                                                                           |
|-----------------------|------------------------------------------------|
| **Property-tax base**                            | Use **last year's assessed value** (no step-up) and apply the current % rate to every future year.                     |
| **Home-maintenance allowance**                   | 1 % of the **previous year's home value** (grows with appreciation).                                                   |
| **Inflation escalators**                         | • Utilities grow at the **same rate you set for CPI (3 % base case)**.<br>• HOA dues rise at the **CPI rate** as well. |
| **Up-front fees** (origination, appraisal, etc.) | **Paid out-of-pocket** at closing; **not** rolled into loan balance.                                                   |
| **MIP / PMI deductibility**                      | Deductible for income-tax purposes; model it as an **interest-type** item on Schedule A.                               |
| **Standard deduction path**                      | Start at **\$15 750 (2025 single)** and **inflate by CPI** (per "Big Beautiful Bill" proposal).                        |
| **Capital-gains rules on sale**                  | **Capital-gain tax if Maya sells** (assume she does at Year 10):                                                       |

• First **\$250 000** of gain (single) or **\$500 000** (married filing jointly) is excluded **if Maya owned & used the home as her principal residence for ≥ 2 of the 5 years before the sale** (IRC § 121).\
• Any remaining gain is taxed at the long-term capital-gain rate (15 % here).\
• Assume she still meets the 2-of-5 test at the Year-10 sale. \|

#### US Tax Assumptions

| Description                                    | Value         |
|------------------------------------------------|---------------|
| Capital-gains exclusion -- **Single**          | **\$250 000** |
| Capital-gains exclusion -- **Married / Joint** | **\$500 000** |
| Tax on non-excluded gain                       | **15 %**      |
| Mortgage-interest deduction limit (principal)  | **\$750 000** |

> **Loan costs are not financed.** Treat appraisal, inspection, and other origination fees as an immediate cash outflow at month 0.
