---
project:
  type: website

website:
  title: "Rent vs Buy – Help Maya Decide"

format:
  html:
    theme: flatly           # keep any Bootstrap theme you like
    css: 
      - styles/latex-fonts.scss   # SCSS / CSS file that loads Latin-Modern

editor: visual
---


## Lab 3 – Buy vs Rent (30-Year Outlook)

::: {.callout-note}
**Watch Maya’s 10-second introduction** before you compare the options.
:::

```{=html}
<!-- 16:9 responsive wrapper -->
<div style="position:relative;padding-bottom:56.25%;height:0;">
  <iframe src="https://www.youtube.com/embed/vzJ31Tz5z-s"
          title="Maya intro"
          style="position:absolute;top:0;left:0;width:100%;height:100%;border:0;"
          allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture"
          allowfullscreen>
  </iframe>
</div>
```

<!-- Example with a Dropbox raw link 
```{=html}
<video controls style="max-width:100%;border-radius:8px;box-shadow:0 2px 6px rgba(0,0,0,.15)">
  <source src="https://youtu.be/vzJ31Tz5z-s" type="video/mp4">
  Sorry, your browser can’t play this video.
</video>
```


```{=html}
<video src="/static/video/maya_intro.mp4"
       controls
       style="max-width:100%;border-radius:8px;box-shadow:0 2px 6px rgba(0,0,0,.15)">
</video>
```
-->


Maya knows Austin housing is pricey, but she also knows staying on the sidelines for too long can be costly. Your job in **Lab 3** is to convert the spreadsheet you started in Lab 2 into a **decision tool** that shows Maya—with one click—how her net worth will look after **10 years** under four realistic paths:

* Buying the **Barton-Springs condo** a stone’s throw from work  
* Buying a smaller **starter home on Lightfoot Trail**  
* Buying a lower-priced **starter in Kyle, TX**, 22 miles away  
* **Continuing to rent** a two-bed South Lamar apartment with a roommate  



At the end you will recommend the path that leaves Maya wealthiest **and** explain the trade-offs of big vs small down-payments, PMI vs FHA mortgage insurance, and lifestyle factors like commute and HOA dues.

---

### 1 Baseline facts you must load into the model
|                             | **Option 1 – Barton-Springs Condo** | **Option 2 – Lightfoot Starter** | **Option 3 – Kyle Starter** | **Rent – South Lamar apt**    |
| --------------------------- | ----------------------------------- | -------------------------------- | --------------------------- | ----------------------------- |
| Price / Monthly rent        | \$649 500                           | \$500 000                        | **\$350 000**               | \$2 295 / mo (≈ \$1 148 each) |
| Beds / Baths / Sq ft        | 2 / 2 / 994                         | 3 / 2 / 1 781                    | 3 / 2 / 1 590               | 2 / 2 / 997                   |
| HOA dues (mo)               | \$485                               | \$125                            | **\$46**                    | —                             |
| Distance to office          | 0.7 mi                              | 4.2 mi                           | **22 mi**                   | 0.9 mi                        |
| **Commute cost (mo)**       | \$0                                 | \$0                              | **\$635**                   | \$0                           |
| Property-tax rate           | 1.8 %                               | 2.0 %                            | **2.05 %**                  | —                             |
| Maintenance allowance       | 1 % of value                        | 1 %                              | 1 %                         | —                             |
| Home-owner’s insurance (mo) | \$100                               | \$110                            | **\$105**                   | —                             |
| Utilities (mo)              | \$180                               | \$220                            | \$220                       | \$180 (split)                 |
| Renter’s insurance (mo)     | —                                   | —                                | —                           | \$15 each                     |
| Home appreciation           | 2.5 % / yr (all)                    |                                  |                             |                               |
| Sale / purchase costs       | 1 % / 3 % (all buys)                |                                  |                             | —                             |

Maya starts on a **$65 000 salary**, expects **3 % raises**, pays **12 % federal tax**, and invests any leftover cash at **8 % pre-tax**. Rent is assumed to rise **3 % per year**. Keep these figures editable but locked on your first submission. The standard deduction for single filers for tax year 2025 is **$15,750**.  The current SALT cap is **$10,000** but test for the updated one under the "**One big beautiful bill Act**" **$40,000**





<!-- ---------- lightweight Zillow-style CSS ---------- -->

```{=html}
<style>
.cards{
  display:grid;
  grid-template-columns:repeat(auto-fill,minmax(320px,1fr));
  gap:1rem;margin-top:1rem
}
.card{
  border:1px solid #ddd;border-radius:8px;box-shadow:0 1px 4px rgba(0,0,0,.06);
  overflow:hidden;background:#fff;font-family:system-ui,sans-serif
}
.card-header{position:relative}
.card-header img{width:100%;height:200px;object-fit:cover}
.badge{
  position:absolute;top:8px;left:8px;padding:2px 6px;border-radius:4px;
  font-size:.75rem;color:#fff;opacity:.9
}
.badge.Option1 { background:#0a7d00; }   /* condo */
.badge.Option2 { background:#006aff; }   /* Lightfoot */
.badge.Option3 { background:#4d4dff; }   /* Kyle */
.badge.rentOption { background:#ff8c00; }


.card-body{padding:.75rem}
.price{margin:0;font-size:1.25rem;font-weight:700}
.meta,.address{margin:.2rem 0;color:#555;font-size:.9rem}
.btn{
  display:inline-block;margin-top:.3rem;padding:4px 8px;font-size:.8rem;
  color:#006aff;border:1px solid currentColor;border-radius:4px;text-decoration:none
}
.btn:hover{background:#f0f8ff}
</style>
```
::: callout-note
These three properties have been pre-selected for Maya (distance to 2222 Barton Springs Rd and price range fit).\
Click **"View Listing"** to open the live Zillow page.  **Please note that prices may have changed since the time of recording this assignment.**
:::

## Buying

```{r cards, echo=FALSE, results='asis', warning=FALSE}
library(jsonlite)
library(htmltools)

# --- read the JSON once ---
listings <- fromJSON("data/maya_listings.json", simplifyDataFrame = FALSE)

# --- helpers -------------------------------------------------
badge_html <- function(category) {
  labs <- c(Option1 = "Option 1",
            Option2 = "Option 2",
            Option3 = "Option 3")
  div(class = paste("badge", category), labs[[category]])
}

one_card <- function(h) {
  div(class = "card",
      div(class = "card-header",
          img(src = h$image_url, alt = "", loading = "lazy"),
          badge_html(h$category)
      ),
      div(class = "card-body",
          p(class = "price", sprintf("$%s", format(h$price, big.mark=","))),
          p(class = "meta",
            sprintf("%s bd | %s ba | %s sq ft",
                    h$beds, h$baths, format(h$sqft, big.mark=","))),
          p(class = "meta",
            sprintf("HOA $%s/mo • %.2f mi to office",
                    format(h$hoa_monthly, big.mark=","), h$distance_miles)),
          p(class = "address", h$address),
          a(class = "btn", href = h$listing_url, target = "_blank",
            "View Listing ↗")
      )
  )
}

# --- build the grid of cards and return it as HTML ------------
tagList(
  div(class = "cards", lapply(listings, one_card))
) |> browsable()
```

## Renting

```{r rent_card, echo=FALSE, results='asis', warning=FALSE}
library(jsonlite)
library(htmltools)

rent <- fromJSON("data/maya_rent.json", simplifyDataFrame = FALSE)

rent_card <- function(h){
  div(class = "card",
    div(class = "card-header",
        img(src = h$image_url, alt = "", loading = "lazy"),
        div(class = "badge rentOption", "Rent Option")
    ),
    div(class = "card-body",
        p(class = "price",
          sprintf("$%s / mo  (%s each)",
                  format(h$rent_total, big.mark = ","),
                  format(h$rent_per_person, big.mark = ","))),
        p(class = "meta",
          sprintf("%s bd | %s ba | %s sq ft",
                  h$beds, h$baths, format(h$sqft, big.mark=","))),
        p(class = "meta",
          sprintf("%.2f mi to office", h$distance_miles)),
        p(class = "address", h$address),
        a(class = "btn", href = h$listing_url, target = "_blank",
          "View Listing ↗")
    )
  )
}

tagList(div(class = "cards", rent_card(rent))) |> browsable()

```


---

### 2 Financing menus you must include

To make the tool flexible, create two dropdowns on your Inputs sheet:

1. **Financing Type** → Conventional | FHA | Rent  
2. **Down-payment %** → 20 | 5 | 3.5  *(ignored if “Rent” is chosen)*  

Your formulas must react instantly to these choices. The table below defines the default rates and insurance rules; students may test other down-payments, but the memo must cite the scenarios in **bold**.
<!--
| Path                       | Default DP | Rate | Up-front fee | Monthly insurance logic |
|----------------------------|-----------:|------|--------------|-------------------------|
|Conventional **20 %**       | 20 % | 6.50 % | none | none |
|Conventional **5 % + PMI**  | 5 %  | 6.60 % | none | PMI = 0.65 %/yr of balance until **LTV ≤ 78 %** |
|FHA **3.5 %**               | 3.5 %| 6.25 % | 1.75 % upfront (financed) | MIP = 0.55 %/yr; lasts 30 yrs when DP \< 10 % |
|Rent                        | —   | —      | —    | Renter’s insurance $15/mo |
-->

| Path | Down-payment selector | Rate (30-yr fixed) | Up-front fee | Monthly insurance rule |
|------|----------------------|--------------------|--------------|------------------------|
| **Conventional – 20 % down** (baseline) | 20 % **(editable)** | **6.50 %** | none | **none** |
| **Conventional – Low-down + PMI** | 3 % – 10 % **(student chooses)**<br>*use 5 % in the memo unless you test others* | **6.60 %** *(⅛ pt hit for low-down)* | none | **PMI = 0.65 % / yr × loan balance**<br>• Paid monthly<br>• Auto-cancels at **78 % LTV** (borrower may request at 80 %) |
| **FHA – High** | > 10%  % **(editable; may test 20 % & 15 %)** | **6.15 %** | **1.75 %** of loan, financed | **MIP = 0.50  % / yr × loan balance drops after 11 yrs** or when LTV ≤ 78 % |
| **FHA – Low-down** | 3.5 % **(editable; may test 5 % & 10 %)** | **6.25 %** | **1.75 %** of loan, financed | **MIP = 0.55 % / yr × loan balance**<br>• Paid monthly<br>• If original DP \< 10 %, lasts **full 10 yrs**<br>|

<!--
---

### 3 What to build (methodical steps)

**a.  Workbook skeleton.**  could include : *Inputs*, *Amort & Cash-flow* (120 monthly rows), *Equity & Investment*, and *Dashboard*.  
**b.  Core engine.**  Use `PMT()` for  Principal and Interest (P&I).  Add PMI or MIP each month based on the rules above and turn it off automatically when the loan-to-value condition is met.  Feed net cash flow to Maya’s investment column.  
**c.  10-year summary.**  After 120 cash flows compute sale proceeds, subtract remaining loan, add portfolio balance, and report **net worth**.  
**d.  Self-check.**  For Option 2 with Conventional 20 % down, your monthly Principal and Interest (P&I) should read ≈ $2528.  
**e.  Dashboard & memo.**  Show a table and bar chart ranking: Conventional 20 %, Conventional 5 %, FHA 3.5 %, Rent.  In ≤ one page explain which wins, why, and how PMI or MIP shifted the result.
-->

---

### 3  Workbook blueprint & milestones

#### A Sheet architecture (4 tabs)

| Sheet | Purpose | Minimum structure & naming |
|-------|---------|----------------------------|
|**Inputs** *(green cells only)* | One-stop control panel | • **Property selector** → Option 1 / 2 / 3 / Rent (data-validated list)<br>• **Financing Type** → Conventional / FHA / Rent<br>• **Down-payment %** → 20 / 5 / 3.5 *(greyed out for Rent)*<br>• Baseline assumptions table (**blue text**, locked) – price, tax rate, HOA, utilities, inflation, salary, returns, etc.<br>• Named ranges e.g. `selProp`, `finType`, `dpct`, `price`, `hoa`, `cpi`. |
|**Amort-Cashflow** *(120 rows, one per month)* | Monthly engine | Columns:<br>`Month` • `BeginBal` • `Interest` • `Principal` • `EndBal` • `PMI_MIP` • `Taxes` • `HOA` • `Maint` • `Insur` • `Utilities` • `Commute` • **`NetCashFlow`** (outgo when buying, surplus when renting) |
|**Equity-Investment** | Year-end snapshot | For each year 0-10:<br>`HomeValue` • `LoanBal` • `Equity` • `Portfolio` • **`NetWorth = Equity + Portfolio`** |
|**Dashboard** | Presentation layer | • 4×4 table of 10-yr Net Worth (Conv 20, Conv 5, FHA 3.5, Rent)<br>• Bar/column chart bound to that table<br>• Conditional format: green fill on the max value |

---

#### B Core engine formulas (Amort-Cashflow)

1. **P&I** `=IF(finType="Rent",0, PMT(rate/12, 360, -LoanAmt))`  
2. **PMI (Conventional)** `=IF(AND(finType="Conventional", dpct<0.20, LTV>0.80), LoanBal*0.0065/12, 0)`  
3. **MIP (FHA)** `=IF(finType="FHA", LoanBal*mipRate/12, 0)` with `mipRate` chosen from table (0.55 % or 0.50 %).  
4. **Taxes / HOA / Utilities / Maint / Commute** link to named rates and apply CPI escalator each anniversary.  
5. **NetCashFlow** sum of all outflows; *if Rent*, flip sign and add to **`InvestmentContribution`** on Equity-Investment sheet.

---

#### C Ten-year roll-up (Equity-Investment)

*Row “10” calculations*

---

#### Additional modelling assumptions  

| Item | Value / Rule |
|------|--------------|
|**Property-tax base**| Use **last year’s assessed value** (no step-up) and apply the current % rate to every future year. |
|**Home-maintenance allowance**| 1 % of the **previous year’s home value** (grows with appreciation). |
|**Inflation escalators**| • Utilities grow at the **same rate you set for CPI (3 % base case)**.<br>• HOA dues rise at the **CPI rate** as well. |
|**Up-front fees** (origination, appraisal, etc.)| **Paid out-of-pocket** at closing; **not** rolled into loan balance. |
|**MIP / PMI deductibility**| Deductible for income-tax purposes; model it as an **interest-type** item on Schedule A. |
|**Standard deduction path**| Start at **\$15 750 (2025 single)** and **inflate by CPI** (per “Big Beautiful Bill” proposal). |
|**Capital-gains rules on sale**|  **Capital-gain tax if Maya sells** (assume she does at Year 10):  
  • First **$250 000** of gain (single) or **$500 000** (married filing jointly) is excluded **if Maya owned & used the home as her principal residence for ≥ 2 of the 5 years before the sale** (IRC § 121).  
  • Any remaining gain is taxed at the long-term capital-gain rate (15 % here).  
  • Assume she still meets the 2-of-5 test at the Year-10 sale. |

##### US Tax Assumptions

| Description | Value |
|-------------|-------|
|Capital-gains exclusion – **Single** | **\$250 000** |
|Capital-gains exclusion – **Married / Joint** | **\$500 000** |
|Tax on non-excluded gain | **15 %** |
|Mortgage-interest deduction limit (principal) | **\$750 000** |

> **Loan costs are not financed.**  Treat appraisal, inspection, and other origination fees as an immediate cash outflow at month 0.


### 4 Deliverables & grading

Upload **one Excel workbook** and **one-page PDF memo** to Canvas by **Tuesday July 29, 11:59 pm**, No Late work .  
Grading: 30 pts correct model • 30 pts accurate 10-yr net-worth calcs • 25 pts memo clarity • 15 pts discussion of PMI / MIP vs risk factors.  Late work loses 10 % per day.


| Item | Pts |
|------|-----|
|Spreadsheet updated with all inputs & working formulas | **30** |
|Correct 30-year net-worth figures (±\$500) | **30** |
|Memo: clear recommendation + numeric support | **25** |
|Discussion of FHA impact & qualitative factors | **15** |

**Build it as if Maya will use it tomorrow—clarity first, shortcuts last.** 

*Good luck!*


### Notes 



